### 2024-01-16 mss.tex
first section

\documentclass[twoside]{article}

\usepackage{ustj}

%\newcommand{\authorname}{}
\newcommand{\authorpatp}{\patp{mopfel-winrux}}
\newcommand{\affiliation}{Native Planet}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{What Hast Earth to do With Mars}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{What Hast Earth to do With Mars: An explanation of Urbit's IPC interfaces}
\author{~\authorpatp \\ \affiliation}
\date{}

\begin{document}

\maketitle
\thispagestyle{firststyle}

\begin{abstract}
In this groundbreaking paper, we embark on a cosmic journey with the renowned character Ernest P. Worrell as he ventures into the unexplored realm of Martian computing. Drawing inspiration from Ernest's comically ingenious encounters with everyday challenges, we investigate the foundations of what we term "Artificial Stupidity." As Ernest grapples with Martian technology, we delve into the intricacies of programming errors, algorithmic missteps, and the curious phenomena that arise when human-like intelligence meets extraterrestrial computing systems. Our analysis sheds light on the unexpected intersections between humor, artificial intelligence, and the cosmic absurdity of Martian software. Join us in this interplanetary exploration as we unravel the mysteries of Artificial Stupidity through the lens of Ernest's interstellar escapades.
\end{abstract}

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{1}

\tableofcontents

\section{Introduction}

In the realm of software, two worlds exist in stark contrast: 
Earth, the familiar domain of established software paradigms, bustling with complex and varied digital ecosystems, 
and Mars, the enigmatic realm of Urbit, a system as pristine and archaic as the Martian landscape itself. 
This article sets forth on an expedition to unravel the mysteries of this Martian software landscape, 
exploring how the intricate processes of terrestrial technology can establish a dialogue with the Martian code.

Unlike Earth, where software is a tapestry of evolution and patchwork, Mars presents a realm of purity. 
It is a world where software exists as a Maxwellian construct, untouched and unaltered by the non-Maxwellian intricacies that characterize Earth’s digital environments. 
Yet, in this isolation lies a profound awareness - an understanding by the Martians that there exist diverse forms of computing with which they must engage. 
Mars, in its wisdom, accommodates these external entities, allowing them to interact in a dialect familiar to its inhabitants: the language of nouns.

Our journey into this uncharted territory begins with a deep dive into the architecture of Urbit, focusing on the foundational layer upon which Arvo, Urbit's kernel, operates.
This exploration will illuminate the operational dynamics and philosophical underpinnings that make Urbit a universe apart. 
Following this, we delve into the intricacies of two pivotal components of Urbit’s ecosystem: 
\texttt{\%khan} and \texttt{\%lick}. 
These vanes serve as distinct conduits between the terrestrial and the Martian, facilitating not just interaction but a nuanced form of control and collaboration with Arvo.
Through this exploration, we aim to demystify the enigma of Urbit and bridge the cosmic gap between Earth’s established software paradigms and the Martian vision of computational purity and simplicity.

\section{Urbit's Unique Landscape}

\subsection{Arvo: The Heart of Martian Computing}
Arvo, also known as Urbit OS, represents a paradigm shift in operating system design. Unlike traditional operating systems that operate on preemptive multitasking and complex event networks, Arvo is a purely functional operating system, characterized by its deterministic nature and compact size. The entire Urbit stack is about 30,000 lines of code, with Arvo itself being only around 1,000 lines. This small codebase is intentional, reflecting a philosophy that system administration complexity is directly proportional to code size.

Arvo's design philosophy positions it not just as an operating system but as a new frontier in the peer-to-peer internet space. Unlike traditional operating systems like Windows, macOS, or Linux, Arvo doesn't aim to replace them; instead, it offers a unique user experience akin to a web browser, yet it is a fully-fledged OS in its own right. This distinctive approach enables Arvo to function within a virtual machine, theoretically capable of running on bare metal.

Arvo's unique architecture avoids what is referred to as "event spaghetti" by maintaining a clear causal chain for every computation. Each chain begins with a Unix I/O event and progresses through a series of steps until the computation's terminal cause. This deterministic nature allows for a high level of predictability and control, a stark contrast to the non-deterministic nature of most Earth-based operating systems.

The kernel's design as a "purely functional operating system" – or more accurately, an "operating function" – underscores its uniqueness. Arvo operates on the principle that the current state is a pure function of its event log, a record of every action ever performed. This determinism is a significant deviation from the norm, where operating systems allow for programmatic alteration of global variables affecting other programs.

Arvo handles nondeterminism in an innovative way. The system, in essence, behaves like a stateful packet transceiver – events are processed, but there's no guarantee of completion, mirroring the behavior of packet dropping in networking. This approach to handling nondeterminism is unique and aligns with Arvo's overall philosophy of simplicity and determinism.

Arvo's determinism is a key feature, stacking it atop a frozen instruction set known as Nock. This concept, though new to operating systems, is not foreign to computing. Similar to how CPU instruction sets like x86-64 are frozen at the chip level, Arvo freezes the instruction set at a higher level, enabling deterministic computation. This high-level determinism contrasts sharply with the non-determinism typically found in Earth's operating systems.

Handling nondeterminism in Arvo is akin to a heuristic decision-making process similar to dropping a packet in networking. This approach views Arvo as a stateful packet transceiver, where the completion of events is not guaranteed, a stark difference from traditional computing models. Additionally, because Arvo runs on a VM, it can obtain nondeterministic information, such as stack traces from infinite loops, through the interpreter beneath it, further enhancing its operational capabilities.

The vision of Urbit, with Arvo at its core, is to transition from developer-hosted web services on multiple foreign servers to self-hosted applications on a personal server. Each architectural decision in Arvo aligns with this vision, emphasizing user ownership and management of data. This approach sets it apart from the conventional model of cloud-based services and centralized data management prevalent in Earth-based systems.

\subsection{Vere: The Substrate of Martian Technology}

Vere, the runtime of Urbit, plays a critical role in actualizing the Martian computing model on Earth-based hardware. As the Nock interpreter written in C, Vere is intricately optimized to run Arvo, ensuring seamless translation of its deterministic operations into practical execution on conventional systems.

Key to Vere's functionality is how it processes events. Events from the Unix environment are passed to Vere, which then translates and injects them into Arvo. This mechanism demonstrates a profound integration between Vere and Arvo, where Vere possesses direct knowledge of Arvo's state. Notably, Vere can 'scry' or query Arvo's state without altering it, retrieving information as needed. This capability allows Vere to maintain the integrity of Arvo's deterministic model while enabling dynamic interaction with the outside world.

Urbit's use of Unix as its BIOS is a pivotal aspect of its runtime operation. By leveraging the robust, widely-used Unix system, Vere ensures that Urbit can run on a broad range of hardware platforms, effectively making Urbit's Martian technology accessible and operable in Earth's diverse computing environments. This strategic use of Unix not only provides the necessary I/O and optimizations for Arvo but also ensures that Urbit's advanced and unique software architecture can function in tandem with the existing, established hardware and software ecosystems of Earth.

Moreover, Vere's design allows it to inject events into Arvo, a feature crucial for maintaining the fluid communication and interaction between the Urbit system and its underlying hardware and software infrastructure. This bidirectional communication channel ensures that Arvo can respond to external stimuli while remaining true to its functional and deterministic nature.

Thus, Vere serves as more than just a bridge between Martian technology and Earthly hardware; it is a finely tuned conduit, optimized to uphold and facilitate the unique computational paradigm that Arvo embodies.

\section{Background and Literature}

Exposite the relevant background and literature.

\begin{enumerate}
  \item  Prefer to \texttt{enumerate}.
  \item  It is easier to refer to than \texttt{itemize}.
\end{enumerate}

\noindent
Mark \texttt{\textbackslash noindent} in paragraphs that continue the thought of an \texttt{\textbackslash enumerate} block.

\section{Urbit's Implementation}

Oftentimes you will then turn to exploring how Urbit or closely related systems like Nockchain have approached or considered a problem.

\lstset{language=C}
\begin{lstlisting}
#include <stdio.h>

double compute(float a, float b) {
    return a * a + b * b;
}

int main() {
    float x = 2.5;
    float y = 3.7;

    double result = compute(x, y);

    printf("Result: %lf\n", result);

    return 0;
}
\end{lstlisting}

\noindent
That code snippet was part of the text, and not a standalone entity to which we make separate repeated reference.

\begin{figure}
  \begin{lstlisting}[language=Python, caption={Example Python Code}, label={lst:example}]
  def hello_world():
      print("Hello, world!")
  \end{lstlisting}
\end{figure}

Refer to the code listing: Listing~\ref{lst:example}.  (Note that we don't suppress indentation since it's a float.)

\section{A Wild Ernest Appears}

You can have as many sections as make sense.  Only sections and subsections appear in the table of contents.

\section{Conclusion}

To summarize, why did you write it?  Why do we care?  What impact should it have on Urbit development?

Your bibliography is a separate BibTeX file.  We use the \texttt{plainnat} bibliography style.  You can use \texttt{natbib} citation commands like \texttt{\textbackslash citep\{wikipedia\}} for parenthesized references.  Use \texttt{\textbackslash citet\{wikipedia\}} for inline references.  You can also use \texttt{\textbackslash citeauthor\{wikipedia\}} for the principal author's name.

``You can use traditional TeX---or LaTeX---representations'' \citep{Varney1987}.

“Or you can use fancy quotes—and symbols.”

\bibliographystyle{plainnat}
\bibliography{ernest}
\end{document}

### 2024-01-26 mss.tex
added sections

@@
 
-\section{Background and Literature}
+\section{Khan - The High-Level Thread Interface of Urbit}
 
-Exposite the relevant background and literature.
+Khan, as a crucial component of Urbit's architecture, functions as a high-level thread interface, pivotal for managing both internal and external communications within the Urbit ecosystem.
+It is designed to efficiently handle complex I/O operations, employing an IO monad coupled with an exception monad.
+This dual-mechanism approach allows Khan to adeptly manage complex IO tasks while effectively handling potential failures.
+
+Khan's role in Urbit can be likened to that of a control plane.
+Its primary function is to execute threads via a Unix Socket and relay the results back.
+This setup supports various thread-running modes, enhancing the flexibility of Urbit's computing environment and allowing for diverse interaction methods with both internal and external processes.
+
+At its core, Khan's design revolves around the use of 'threads', which are monadic functions designed to handle arguments and produce results.
+These threads are optimized for I/O operations, leveraging the IO monad for structured and efficient task handling.
+The exception monad further ensures robust management of potential failures and unexpected events.
+
+Khan's implementation ensures that Urbit serves as a personal server, replacing multiple developer-hosted web services on foreign servers with self-hosted applications on a single personal server.
+Its design underscores Urbit's commitment to user ownership and data management, setting it apart from the conventional model of cloud-based services and centralized data management prevalent in Earth-based systems.
+
+\section{\texttt{conn.c}} %Rewrite so its not a blatant copy of alex's work
+
+\texttt{conn.c}\footnote{\url{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}} is a driver in Vere.
+It is a part of the "King" (a.k.a.
+"Urth") process.
+It exposes a \texttt{Unix domain socket}\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}} at \texttt{/path/to/pier/.urb/conn.sock} for sending/receiving data from external processes.
+
+The functionality of conn.c, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of urbit's control plane.
+It serves as a way for unix processes to receive insights about what is happening on urbit.
+It also allows for pass commands either \%khan and the injection of raw kernel moves into the system.
+The implication of this means that whoever is running your urbit runtime controls your urbit.
+
+From a technical perspective, conn.c accepts newt-encoded \texttt{++jam}med nouns which take the form \texttt{[request-id command arguments]}.
+This structure allows for a variety of commands to be executed, including
 
 \begin{enumerate}
-  \item  Prefer to \texttt{enumerate}.
-  \item  It is easier to refer to than \texttt{itemize}.
+  \item \texttt{\%ovum} - the injection of raw kernel moves 
+  \item \texttt{\%fyrd} - a direct shortcut to Khan commands
+  \item \texttt{\%urth} - runtime subcommands like \texttt{\%pack}` or \texttt{\%meld}
+  \item \texttt{\%peek} - namespace scry requests into Arvo
+  \item \texttt{\%peel} - emulated namespace scry requests into Vere
 \end{enumerate}
 
-\noindent
-Mark \texttt{\textbackslash noindent} in paragraphs that continue the thought of an \texttt{\textbackslash enumerate} block.
+A valid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[request-id output]}, where:
+\begin{itemize}
+  \item \texttt{request-id} matches the input \texttt{request-id}
+  \item \texttt{output} depends on the \texttt{command}
+\end{itemize}
 
-\section{Urbit's Implementation}
+An invalid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[0 \%bail error-code error-string]}.
 
-Oftentimes you will then turn to exploring how Urbit or closely related systems like Nockchain have approached or considered a problem.
+\subsection{\%ovum}
 
-\lstset{language=C}
-\begin{lstlisting}
-#include <stdio.h>
+The argument to an \texttt{\%ovum} command is a raw kernel move which is injected directly into the Arvo event loop.
+This is a very powerful - and potentially dangerous - tool.
+For example, if a ship somehow got into a state where Clay was no longer working properly (meaning new files could not be compiled to fix the state of the kernel), the source code for a new, working Clay could be directly injected into the ship using an \texttt{\%ovum}.
 
-double compute(float a, float b) {
-    return a * a + b * b;
-}
+The output of an \texttt{\%ovum} command is:
+\begin{itemize}
+  \item \texttt{[\%news \%done]} if the move completed successfully
+  \item \texttt{[\%news \%drop]} if the move was dropped
+  \item \texttt{[\%bail goof]} if an error occurred
+\end{itemize}
 
-int main() {
-    float x = 2.5;
-    float y = 3.7;
+\subsection{\%fyrd}
 
-    double result = compute(x, y);
+\texttt{\%fyrd} is a direct shortcut to the Khan vane. The arguments to a \texttt{\%fyrd} command are (in order):
+\begin{enumerate}
+  \item The name of the desk in which the thread lives (e.g. \texttt{\%base}) or \texttt{beak} for the thread (e.g. \texttt{[\%zod \%base \%10]})
+  \item The name of the thread (e.g. \texttt{\%hi})
+  \item Mark to which the output should be cast (e.g. \texttt{\%tape})
+  \item Mark for how to interpret the input argument to thread (e.g. \texttt{\%ship})
+  \item Input argument to thread (e.g. \texttt{\~zod})
+\end{enumerate}
 
-    printf("Result: %lf\n", result);
+The output of a \texttt{\%fyrd} command is \texttt{[\%avow (each page goof)]}, the value of \texttt{each} depending on whether the thread succeeded or not.
 
-    return 0;
-}
-\end{lstlisting}
+\subsection{\%urth}
 
-\noindent
-That code snippet was part of the text, and not a standalone entity to which we make separate repeated reference.
+The argument to the \texttt{\%urth} command is a subcommand for the action to perform. Currently, the only valid commands are \texttt{\%pack} and \texttt{\%meld}.
 
-\begin{figure}
-  \begin{lstlisting}[language=Python, caption={Example Python Code}, label={lst:example}]
-  def hello_world():
-      print("Hello, world!")
-  \end{lstlisting}
-\end{figure}
+\texttt{\%urth} will return \texttt{\%\&} if given a valid command as input, otherwise it will return \texttt{[0 \%bail 0xfffffff9 \%urth-bad]}. No other output is emitted.
 
-Refer to the code listing: Listing~\ref{lst:example}.  (Note that we don't suppress indentation since it's a float.)
+\subsection{\%peek}
 
-\section{A Wild Ernest Appears}
+The \texttt{\%peek} command is used to perform a namespace read request (a.k.a.
+scry) using Arvo's external peek interface \texttt{arm +22 in arvo.hoon}\footnote{\url{https://github.com/urbit/urbit/blob/develop/pkg/arvo/sys/arvo.hoon#L1774}}.
+The argument to \texttt{\%peek} is the \texttt{nom} input to \texttt{+peek} in \texttt{arvo.hoon} (\texttt{lyc} is auto-filled as \texttt{[~ ~]}, i.e., "request from self").
+That is to say that the argument to \texttt{\%peek} must have type:
+\begin{verbatim}
+$+  each  path
+$%  [%once vis=view syd=desk tyl=spur]
+    [%beam vis=view bem=beam]
+==
+\end{verbatim}
+Practically speaking, this means that the input will look like one of these three examples:
+\begin{verbatim}
+[%& p=path]
+[%| p=[%once vis=view syd=desk tyl=spur]]
+[%| p=[%beam vis=view bem=beam]]
+\end{verbatim}
+Where:
+\begin{itemize}
+  \item \texttt{path} is a \texttt{[view beam]}, with the \texttt{view} passed in as a \texttt{coin}
+  \item \texttt{view} is the vane code for the scry, as well as an optional care, possibly appended to the vane (e.g. \texttt{\%j}, \texttt{\%gx}, etc.)
+  \item \texttt{beam} is a \texttt{[beak spur]}
+  \item \texttt{desk} is used to auto-generate a \texttt{beak}: \texttt{[our desk now]}
+  \item \texttt{spur} is the scry endpoint for the agent or vane
+\end{itemize}
 
-You can have as many sections as make sense.  Only sections and subsections appear in the table of contents.
+The output of a \texttt{\%peek} command is \texttt{[\%peek (unit (unit scry-output))]}, where \texttt{~} means that the scry endpoint is invalid, and \texttt{[~ ~]} means that the scry resolved to nothing.
 
+\subsection{\%peel}
+
+\texttt{\%peel} attempts to emulate a scry-like namespace, like the one used by Arvo and accessed by \texttt{\%peek}.
+The argument to \texttt{\%peel} should be a path.
+Valid paths result in a non-null \texttt{unit} containing the result of the scry.
+Invalid paths result in null (i.e.
+\texttt{~}).
+The valid paths and the data they return are:
+
+\begin{verbatim}
+/help   (unit (list path))  Supported %peel paths
+/live   (unit ~)            Pier health check; succeeds if pier is running
+/khan   (unit ~)            Khan health check; succeeds if Khan vane is running
+/info   (unit mass)         Pier info as a mass
+/v      (unit @t)           Returns version of the Vere binary as a cord
+/who    (unit @)            Returns the Azimuth identity of the ship as an atom
+\end{verbatim}
+
+Note that the pier info above is returned as a \texttt{mass} report, i.e.
+type \texttt{(pair cord (each * (list mass)))}.
+This is not the same as the \texttt{|mass} memory report.
+\texttt{/mass} is meant to be a valid \texttt{\%peel} path which returns the \texttt{|mass} memory report, but it is currently unimplemented.
+
+\section{\%lick}
+
+Altho also dealing with interprocess communication, Lick was designed for a very different scenario than Khan:  to allow external processes, in particular hardware drivers, to intercommunicate with Urbit.
+(This breached the Earth/Mars divide.)  Thus \texttt{/sys/vane/lick} focuses on instrumenting a low-level noun interfaces over domain sockets.
+
+\texttt{\%lick} manages IPC ports, and the communication between Urbit applications and POSIX applications via these ports.
+Other vanes and applications ask \texttt{\%lick} to open an IPC port, notify it when something is connected or disconnected, and transfer data between itself and the Unix application.
+
+Lick works by opening a Unix socket for a particular process, which allows serialized IPC communications.
+These involve a jammed noun so the receiving process needs to know how to communicate in nouns.
+
+The IPC ports Lick creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of the \texttt{SOCK\_STREAM} type.
+
+The connexions are made via filepaths in \texttt{.urb/dev} of the pier.
+
+The format is:
+
+\begin{verbatim}
+V.BBBB.JJJJ.JJJJ...
+\end{verbatim}
+
+\begin{itemize}
+  \item \texttt{V} is the version
+  \item \texttt{B} is the jam size in bytes (little endian)
+  \item \texttt{J} is the jammed noun (little endian)
+\end{itemize}
+
+The process on the host OS must therefore strip the first 5 bytes, \texttt{++cue}\footnote{\url{https://docs.urbit.org/language/hoon/reference/stdlib/2p#cue}} the jamfile, check the mark and (most likely) convert the noun into a native data structure.
+
+\subsection{\texttt{/sys/lull} Definition}
+
+\begin{verbatim}
+::                                              ::::
+::::                    ++lick                    ::  (1j) IPC
+  ::                                            ::::
+++  lick  ^?
+  |%
+  +$  gift                                      ::  out result <-$
+    $%  [%spin =name]                           ::  open an IPC port
+        [%shut =name]                           ::  close an IPC port
+        [%spit =name =mark =noun]               ::  spit a noun to the IPC port
+        [%soak =name =mark =noun]               ::  soak a noun from the IPC port
+    ==
+  +$  task                                      ::  in request ->$
+    $~  [%vega ~]                               ::
+    $%  $>(%born vane-task)                     ::  new unix process
+        $>(%trim vane-task)                     ::  trim state
+        $>(%vega vane-task)                     ::  report upgrade
+        [%spin =name]                           ::  open an IPC port
+        [%shut =name]                           ::  close an IPC port
+        [%spit =name =mark =noun]               ::  spit a noun to the IPC port
+        [%soak =name =mark =noun]               ::  soak a noun from the IPC port
+    ==
+  ::
+  +$  name  path
+  --  ::lick
+\end{verbatim}
+
+To evaluate what \texttt{/sys/vane/lick} is doing, we need to look at Unix's IPC model briefly.
+IPC ("interprocess communication") describes any way that two processes in an operating system's shared context have to communicate with each other.
+\texttt{\%lick} focuses on Unix domain sockets\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}}, which are just communication endpoints\footnote{\url{https://man7.org/linux/man-pages/man7/unix.7.html}}.
+
+For instance, a valid use of \texttt{\%lick} would use cards that look like this:
+
+\begin{verbatim}
+++  init  [[%pass / %arvo %l %spin /control]~ this]
+::
+++  on-arvo
+  |=  [=wire =sign-arvo]
+  ?+  sign-arvo  (on-arvo:def wire sign-arvo)
+      [%lick %soak *]
+      ?+  mark.sign-arvo  [~ this]
+      ::
+        %connect
+      ~&  >  "connect"
+      :_  this  [%pass /spit %arvo %l %spit /control %init area.state]~
+  ==  ==
+::
+++  send-state
+  |=  =state
+  ^-  card:agent:gall
+  [%pass /spit %arvo %l %spit /control %state [slick:state face.state food.state live.state]]
+\end{verbatim}
+
+The vane definition of \texttt{/sys/vane/lick} is even simpler than \texttt{/sys/vane/khan}:  it has no \texttt{++abet} core and primarily communicates to the \texttt{unix-duct} in its state.
+The \texttt{owner} is a \texttt{duct} to handle the return \texttt{\%soak}.
+
+Gall needs to wrap \texttt{\%soak} and \texttt{\%spit} to route properly.
+See e.g.
+\texttt{\%++ap-generic-take}.
+This lets multiple agents share sockets with the same name, and each agent can have its own folder.
+
+%\section{Background and Literature}
+
+%Exposite the relevant background and literature.
+
+%\begin{enumerate}
+  %\item  Prefer to \texttt{enumerate}.
+  %\item  It is easier to refer to than \texttt{itemize}.
+%\end{enumerate}
+
+%\noindent
+%Mark \texttt{\textbackslash noindent} in paragraphs that continue the thought of an \texttt{\textbackslash enumerate} block.
+%
+%\section{Urbit's Implementation}
+%
+%Oftentimes you will then turn to exploring how Urbit or closely related systems like Nockchain have approached or considered a problem.
+%
+%\lstset{language=C}
+%\begin{lstlisting}
+%#include <stdio.h>
+%
+%double compute(float a, float b) {
+    %return a * a + b * b;
+%}
+%
+%int main() {
+    %float x = 2.5;
+    %float y = 3.7;
+%
+    %double result = compute(x, y);
+%
+    %printf("Result: %lf\n", result);
+%
+    %return 0;
+%}
+%\end{lstlisting}
+%
+%\noindent
+%That code snippet was part of the text, and not a standalone entity to which we make separate repeated reference.
+%
+%\begin{figure}
+  %\begin{lstlisting}[language=Python, caption={Example Python Code}, label={lst:example}]
+  %def hello_world():
+      %print("Hello, world!")
+  %\end{lstlisting}
+%\end{figure}
+%
+%Refer to the code listing: Listing~\ref{lst:example}.  (Note that we don't suppress indentation since it's a float.)
+%
+
+%\section{A Wild Ernest Appears}
+%
+%You can have as many sections as make sense.  Only sections and subsections appear in the table of contents.
+%
 \section{Conclusion}

### 2024-02-05 mss.tex
fixed neals comments

--- 2024-01-16 mss.tex
@@
 \section{Introduction}
-
+%Section 1: can do explicit callback to Moron Labs post (we should add that to the standard mss.bib)
+% When did urbit backoff on the earth only calls mars? (eyre and iris) Urbit had to run on a VM
 In the realm of software, two worlds exist in stark contrast: 
@@
 Mars, in its wisdom, accommodates these external entities, allowing them to interact in a dialect familiar to its inhabitants: the language of nouns.
+
+In the original explanation of of Urbit, \citeauthor{Yarvin2010}, explicitly says ``The general structure of cross-planet computation is that Earth always calls Mars; Mars never calls Earth.
+The latter would be quite impossible, since Earth code is non-Maxwellian.
+There is only one way for a Maxwellian computer to run non-Maxwellian code: in a Maxwellian emulator.'' This however has never been true for Urbit as it was implemented.
+Urbit has two vanes \texttt{\%eyre} and \texttt{\%iris} that are written as a specific interface to Earth (a web server and curl, respectively).
+One can even make the argument that Urbit running on a virtual machine specifically breaks this convention.
 
@@
 Unlike traditional operating systems that operate on preemptive multitasking and complex event networks, Arvo is a purely functional operating system, characterized by its deterministic nature and compact size.
-The entire Urbit stack is about 30,000 lines of code, with Arvo itself being only around 1,000 lines.
+The entire Urbit stack is about 80,000 lines of code, with Arvo itself being only around 2,000 lines.
 This small codebase is intentional, reflecting a philosophy that system administration complexity is directly proportional to code size.
-
-Arvo's design philosophy positions it not just as an operating system but as a new frontier in the peer-to-peer internet space.
-Unlike traditional operating systems like Windows, macOS, or Linux, Arvo doesn't aim to replace them; instead, it offers a unique user experience akin to a web browser, yet it is a fully-fledged OS in its own right.
-This distinctive approach enables Arvo to function within a virtual machine, theoretically capable of running on bare metal.
 
@@
 
+The architecture of Vere is bifurcated into two distinct parts: the \texttt{king} and the \texttt{serf}.
+The \texttt{king} component is responsible for handling vane I/O and managing the effects on Earth.
+In contrast, the \texttt{serf} operates as the process running the Arvo virtual machine (VM).
+This division of responsibilities allows for a clear delineation of tasks within the Vere runtime, enhancing its efficiency and robustness.
+
+Communication between the \texttt{king} and \texttt{serf} is achieved through an Inter-Process Communication (IPC) port.
+This IPC mechanism facilitates a continuous dialogue between the two components, ensuring that the \texttt{king} can effectively manage external interactions while the \texttt{serf} maintains the integrity of the Arvo VM.
+
+A core opcode in Urbit's assembly language (\texttt{nock}) is the \texttt{hint} opcode (opcode 11) this allows Urbit to pass hints to Vere which produces effects on Earth.
+This bridges the Earth/Mars gap and allows for vanes to interact outside of Mars.
+
 Key to Vere's functionality is how it processes events.
@@
 
-\section{Background and Literature}
+\section{\%khan - The High-Level Thread Interface of Urbit}
 
-Exposite the relevant background and literature.
+Khan, as a crucial component of Urbit's architecture, functions as a high-level thread interface, pivotal for managing both internal and external communications within the Urbit ecosystem.
+It is designed to efficiently handle complex I/O operations, employing an IO monad coupled with an exception monad.
+This dual-mechanism approach allows Khan to adeptly manage complex IO tasks while effectively handling potential failures.
+
+Khan allows threads to be triggered from outside of Urbit.
+A thread is a monadic function that takes arguments and produces a result.
+It may perform input and output while running, so it is not a pure function.
+A thread's strength is that it can easily perform complex IO operations.
+It uses what's often called the IO monad (plus the exception monad) to provide a natural framework for IO.
+A thread's weakness is that it's impermanent and may fail unexpectedly.
+In most of its intermediate states, it expects only a small number of events (usually one), so if it receives anything it didn't expect, it fails.
+When code is upgraded, it's impossible to upgrade a running thread, so it fails.
+
+Khan's role in Urbit can be likened to that of a control plane.
+Its primary function is to execute threads via a Unix Socket and relay the results back.
+This setup supports various thread-running modes, enhancing the flexibility of Urbit's computing environment and allowing for diverse interaction methods with both internal and external processes.
+
+At its core, Khan's design revolves around the use of 'threads', which are monadic functions designed to handle arguments and produce results.
+These threads are optimized for I/O operations, leveraging the IO monad for structured and efficient task handling.
+The exception monad further ensures robust management of potential failures and unexpected events.
+
+Khan's implementation ensures that Urbit serves as a personal server, replacing multiple developer-hosted web services on foreign servers with self-hosted applications on a single personal server.
+Its design underscores Urbit's commitment to user ownership and data management, setting it apart from the conventional model of cloud-based services and centralized data management prevalent in Earth-based systems.
+
+\section{\texttt{conn.c}}
+
+\texttt{conn.c}\footnote{\url{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}} is a driver in Vere.
+It is a part of the "King" (a.k.a.
+"Urth") process.
+It exposes a \texttt{Unix domain socket}\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}} at \texttt{/path/to/pier/.urb/conn.sock} for sending/receiving data from external processes.
+
+The functionality of conn.c, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of urbit's control plane.
+It serves as a way for unix processes to receive insights about what is happening on urbit.
+It also allows for pass commands either \%khan and the injection of raw kernel moves into the system.
+The implication of this means that whoever is running your urbit runtime controls your urbit.
+
+From a technical perspective, conn.c accepts newt-encoded \texttt{++jam}med nouns which take the form \texttt{[request-id command arguments]}. 
+The newt-encoded format is:
+
+\begin{verbatim}
+V.BBBB.JJJJ.JJJJ...
+\end{verbatim}
+
+\begin{itemize}
+  \item \texttt{V} is the version
+  \item \texttt{B} is the jam size in bytes (little endian)
+  \item \texttt{J} is the jammed noun (little endian)
+\end{itemize}
+
+This structure allows for a variety of commands to be executed, including
 
 \begin{enumerate}
-  \item  Prefer to \texttt{enumerate}.
-  \item  It is easier to refer to than \texttt{itemize}.
+  \item \texttt{\%ovum} - the injection of raw kernel moves 
+  \item \texttt{\%fyrd} - a direct shortcut to Khan commands
+  \item \texttt{\%urth} - runtime subcommands like \texttt{\%pack} or \texttt{\%meld}
+  \item \texttt{\%peek} - namespace scry requests into Arvo
+  \item \texttt{\%peel} - emulated namespace scry requests into Vere
 \end{enumerate}
 
-\noindent
-Mark \texttt{\textbackslash noindent} in paragraphs that continue the thought of an \texttt{\textbackslash enumerate} block.
+A valid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[request-id output]}, where:
+\begin{itemize}
+  \item \texttt{request-id} matches the input \texttt{request-id}
+  \item \texttt{output} depends on the \texttt{command}
+\end{itemize}
 
-\section{Urbit's Implementation}
+An invalid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[0 \%bail error-code error-string]}.
 
-Oftentimes you will then turn to exploring how Urbit or closely related systems like Nockchain have approached or considered a problem.
+\section{\%lick - The Low Level IPC Interface}
 
-\lstset{language=C}
-\begin{lstlisting}
-#include <stdio.h>
+Although also dealing with interprocess communication, \texttt{\%lick} \citep{UIP-101} was designed for a very different scenario than Khan:  to allow external processes, in particular hardware drivers, to intercommunicate with Urbit.
+(This breached the Earth/Mars divide.)  Thus \texttt{/sys/vane/lick} focuses on instrumenting a low-level noun interfaces over domain sockets.
 
-double compute(float a, float b) {
-    return a * a + b * b;
-}
+\texttt{\%lick} manages IPC ports, and the communication between Urbit applications and POSIX applications via these ports.
+Other vanes and applications ask \texttt{\%lick} to open an IPC port, notify it when something is connected or disconnected, and transfer data between itself and the Unix application.
+Lick works by opening a Unix socket for a particular process, which allows serialized IPC communications.
+These involve a jammed noun so the receiving process needs to know how to communicate in nouns.
+The IPC ports Lick creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of the \texttt{SOCK\_STREAM} type.
+The connexions are made via filepaths in \texttt{.urb/dev} of the pier.
 
-int main() {
-    float x = 2.5;
-    float y = 3.7;
+The process on the host OS must therefore strip the first 5 bytes, \texttt{++cue}\footnote{\url{https://docs.urbit.org/language/hoon/reference/stdlib/2p#cue}} the jamfile, check the mark and (most likely) convert the noun into a native data structure.
 
-    double result = compute(x, y);
+To understand what \texttt{\%lick} is doing, we need to look at Unix's IPC model briefly.
+IPC ("interprocess communication") describes any way that two processes in an operating system's shared context have to communicate with each other.
+\texttt{\%lick} focuses on Unix domain sockets\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}}, which are just communication endpoints\footnote{\url{https://man7.org/linux/man-pages/man7/unix.7.html}}.
 
-    printf("Result: %lf\n", result);
+For instance, a valid use of \texttt{\%lick} would use cards that look like this:
 
-    return 0;
-}
-\end{lstlisting}
+\begin{verbatim}
+++  init  [[%pass / %arvo %l %spin /control]~ this]
+::
+++  on-arvo
+  |=  [=wire =sign-arvo]
+  ?+  sign-arvo  (on-arvo:def wire sign-arvo)
+      [%lick %soak *]
+      ?+  mark.sign-arvo  [~ this]
+      ::
+        %connect
+      ~&  >  "connect"
+      :_  this  
+          [%pass /spit %arvo %l %spit /control %init area.state]~
+      ==  
+  ==
+::
+++  send-state
+  |=  =state
+  ^-  card:agent:gall
+  :*  %pass 
+      /spit 
+      %arvo %l 
+      %spit 
+      /control 
+      %state 
+      [slick:state face.state food.state live.state]
+   ==
+\end{verbatim}
 
-\noindent
-That code snippet was part of the text, and not a standalone entity to which we make separate repeated reference.
+The vane definition of \texttt{/sys/vane/lick} is even simpler than \texttt{/sys/vane/khan}:  it has no \texttt{++abet} core and primarily communicates to the \texttt{unix-duct} in its state.
+The \texttt{+\$owner} is a \texttt{+\$duct} to handle the return \texttt{\%soak}.
 
-\begin{figure}
-  \begin{lstlisting}[language=Python, caption={Example Python Code}, label={lst:example}]
-  def hello_world():
-      print("Hello, world!")
-  \end{lstlisting}
-\end{figure}
+Arvo will send three types of \texttt{\%soak}s to an open \texttt{\%lick} port.
+These \texttt{\%soak}s are \texttt{\%connect} when the first connection is established on the Earth-side IPC port, \texttt{\%disconnect} when the last connection is broken on the Earth-side IPC port, and \texttt{\%error} when an error occurs.
+\texttt{\%lick} also will send a \texttt{\%disconnect} \texttt{\%soak} to every agent when Vere is started.
 
-Refer to the code listing: Listing~\ref{lst:example}.  (Note that we don't suppress indentation since it's a float.)
-
-\section{A Wild Ernest Appears}
-
-You can have as many sections as make sense.  Only sections and subsections appear in the table of contents.
+Gall needs to wrap \texttt{\%soak} and \texttt{\%spit} to route properly.
+See e.g.
+\texttt{\%++ap-generic-take}.
+This lets multiple agents share sockets with the same name, and each agent can have its own folder.
 
@@
 
-To summarize, why did you write it?  Why do we care?  What impact should it have on Urbit development?
+As Urbit had to deal with the reality of running on Earth software, Developers had to come up with various ways of communicating between these systems.
+Originally the two vanes \texttt{\%eyre} and \texttt{\%iris} were used as an HTTP client and server, breaking the ``Earth calls Mars; Mars doesn't call Earth'' dynamic.
 
-Your bibliography is a separate BibTeX file.
-We use the \texttt{plainnat} bibliography style.
-You can use \texttt{natbib} citation commands like \texttt{\textbackslash citep\{wikipedia\}} for parenthesized references.
-Use \texttt{\textbackslash citet\{wikipedia\}} for inline references.
-You can also use \texttt{\textbackslash citeauthor\{wikipedia\}} for the principal author's name.
+Recently, two other vanes, \texttt{\%khan} and \texttt{\%lick} have been developed that hold true to the original idea of communication between Earth and Mars.
+These vanes force Earth to speak to Mars in a language it can understand, nouns.
 
-``You can use traditional TeX---or LaTeX---representations'' \citep{Varney1987}.
-
-“Or you can use fancy quotes—and symbols.”
-
-\bibliographystyle{plainnat}
-\bibliography{ernest}
+\printbibliography
 \end{document}

### 2024-02-05 mss.tex
added abstract

@@
 \begin{abstract}
-In this groundbreaking paper, we embark on a cosmic journey with the renowned character Ernest P.
-Worrell as he ventures into the unexplored realm of Martian computing.
-Drawing inspiration from Ernest's comically ingenious encounters with everyday challenges, we investigate the foundations of what we term "Artificial Stupidity." As Ernest grapples with Martian technology, we delve into the intricacies of programming errors, algorithmic missteps, and the curious phenomena that arise when human-like intelligence meets extraterrestrial computing systems.
-Our analysis sheds light on the unexpected intersections between humor, artificial intelligence, and the cosmic absurdity of Martian software.
-Join us in this interplanetary exploration as we unravel the mysteries of Artificial Stupidity through the lens of Ernest's interstellar escapades.
+  This paper explores the innovative approaches undertaken by developers to enable effective communication between the Urbit system and conventional Earth-based software systems.
+Initially, Urbit's integration with external systems relied on the use of two specific vanes, \texttt{\%eyre} and \texttt{\%iris}, serving as HTTP client and server respectively.
+This setup, however, presented challenges to the ideal of autonomous operation envisaged for Urbit, traditionally characterized by the principle that "Earth calls Mars; Mars doesn't call Earth."
+
+  Addressing these challenges, the paper introduces the development of two new vanes, \texttt{\%khan} and \texttt{\%lick}, which embody the core principle of uncontaminated communication.
+These advancements allow for a refined interaction model where Earth-based systems communicate with Urbit in a Urbit's own language, nouns.
+
+The implications of these developments are significant, offering a novel framework for the integration of distinct computational ecosystems.
+This paper discusses the technical and conceptual underpinnings of these vanes, their development process, and the broader implications for Urbit's interaction with conventional software systems.
+
 \end{abstract}
@@
 \section{Introduction}
-%Section 1: can do explicit callback to Moron Labs post (we should add that to the standard mss.bib)
-% When did urbit backoff on the earth only calls mars? (eyre and iris) Urbit had to run on a VM
 In the realm of software, two worlds exist in stark contrast: 
@@
 Originally the two vanes \texttt{\%eyre} and \texttt{\%iris} were used as an HTTP client and server, breaking the ``Earth calls Mars; Mars doesn't call Earth'' dynamic.
-
-Recently, two other vanes, \texttt{\%khan} and \texttt{\%lick} have been developed that hold true to the original idea of communication between Earth and Mars.
+Recently, two other vanes, \texttt{\%khan} and \texttt{\%lick} have been developed that hold true to the necessity idea of communication between Earth and Mars.
 These vanes force Earth to speak to Mars in a language it can understand, nouns.

### 2024-02-23 mss.tex
added abstract and conclusion

--- 2024-02-05 mss.tex
@@
 \begin{abstract}
-In this groundbreaking paper, we embark on a cosmic journey with the renowned character Ernest P.
-Worrell as he ventures into the unexplored realm of Martian computing.
-Drawing inspiration from Ernest's comically ingenious encounters with everyday challenges, we investigate the foundations of what we term "Artificial Stupidity." As Ernest grapples with Martian technology, we delve into the intricacies of programming errors, algorithmic missteps, and the curious phenomena that arise when human-like intelligence meets extraterrestrial computing systems.
-Our analysis sheds light on the unexpected intersections between humor, artificial intelligence, and the cosmic absurdity of Martian software.
-Join us in this interplanetary exploration as we unravel the mysteries of Artificial Stupidity through the lens of Ernest's interstellar escapades.
+In the evolving landscape of digital computation, the interaction between conventional software paradigms and innovative systems like Urbit presents a unique challenge.
+This paper delves into the architectural intricacies and interfaces of Urbit, contrasting its pristine, Maxwellian principles against the complex, evolved nature of Earth's software ecosystems.
+Through a comprehensive examination of Urbit’s kernel (Arvo) and runtime (Vere), alongside its high-level and low-level IPC interfaces (\texttt{\%khan} and \texttt{\%lick}, respectively), we elucidate the mechanisms that facilitate communication and control between these disparate computing realms.
+Our analysis aims to demystify Urbit's architecture and underscore its potential to bridge the gap between the familiar digital environments of Earth and the untapped possibilities of Martian computing.
+
+%  This paper explores the innovative approaches undertaken by developers to enable effective communication between the Urbit system (Mars) and conventional software systems (Earth).
+Initially, Urbit's integration with external systems relied on the use of two specific vanes, \texttt{\%eyre} and \texttt{\%iris}, serving as HTTP client and server respectively.
+This setup, however, presented challenges to the ideal of autonomous operation envisaged for Urbit, traditionally characterized by the principle that ``Earth always calls Mar; Mars never calls Earth.''
+
+%  Addressing these challenges, the paper introduces the development of two new vanes, \texttt{\%khan} and \texttt{\%lick}, which embody the core principle of uncontaminated communication.
+These advancements allow for a refined interaction model where Earth-based systems communicate with Urbit in a Urbit's own language, nouns.
+
+%The implications of these developments are significant, offering a novel framework for the integration of distinct computational ecosystems.
+This paper discusses the technical and conceptual underpinnings of these vanes, their development process, and the broader implications for Urbit's interaction with conventional software systems.
+
 \end{abstract}
@@
 \section{Introduction}
-%Section 1: can do explicit callback to Moron Labs post (we should add that to the standard mss.bib)
-% When did urbit backoff on the earth only calls mars? (eyre and iris) Urbit had to run on a VM
+
 In the realm of software, two worlds exist in stark contrast: 
@@
 and Mars, the enigmatic realm of Urbit, a system as pristine and archaic as the Martian landscape itself. 
-This article sets forth on an expedition to unravel the mysteries of this Martian software landscape, 
-exploring how the intricate processes of terrestrial technology can establish a dialogue with the Martian code.
+This article will explore how Earth software establish a communication with the Martian code.
 
 Unlike Earth, where software is a tapestry of evolution and patchwork, Mars presents a realm of purity. 
-It is a world where software exists as a Maxwellian construct, untouched and unaltered by the non-Maxwellian intricacies that characterize Earth’s digital environments. 
+It is a world where software exists as a Maxwellian construct, it is defined by a small set of axioms \citep{Nock4k}.
+Urbit code is untouched and unaltered by the non-Maxwellian intricacies that characterize Earth’s digital environments.
 Yet, in this isolation lies a profound awareness - an understanding by the Martians that there exist diverse forms of computing with which they must engage. 
-Mars, in its wisdom, accommodates these external entities, allowing them to interact in a dialect familiar to its inhabitants: the language of nouns.
+Mars, accommodates these external entities by allowing them to interact in the Martian language of nouns.
 
-In the original explanation of of Urbit, \citeauthor{Yarvin2010}, explicitly says ``The general structure of cross-planet computation is that Earth always calls Mars; Mars never calls Earth.
+In the original explanation of of Urbit, \citeauthor{Yarvin2010}, explicitly says 
+``The general structure of cross-planet computation is that Earth always calls Mars; Mars never calls Earth.
 The latter would be quite impossible, since Earth code is non-Maxwellian.
-There is only one way for a Maxwellian computer to run non-Maxwellian code: in a Maxwellian emulator.'' This however has never been true for Urbit as it was implemented.
-Urbit has two vanes \texttt{\%eyre} and \texttt{\%iris} that are written as a specific interface to Earth (a web server and curl, respectively).
+There is only one way for a Maxwellian computer to run non-Maxwellian code: in a Maxwellian emulator.''
+This however has never been true for Urbit as it was implemented.
+Urbit has two vanes (best thought of as kernel modules) \texttt{\%eyre} and \texttt{\%iris} that are written as a specific interface to Earth (a web server and client, respectively).
 One can even make the argument that Urbit running on a virtual machine specifically breaks this convention.
 
-Our journey into this uncharted territory begins with a deep dive into the architecture of Urbit, focusing on the foundational layer upon which Arvo, Urbit's kernel, operates.
-This exploration will illuminate the operational dynamics and philosophical underpinnings that make Urbit a universe apart. 
-Following this, we delve into the intricacies of two pivotal components of Urbit’s ecosystem: 
-\texttt{\%khan} and \texttt{\%lick}. 
-These vanes serve as distinct conduits between the terrestrial and the Martian, facilitating not just interaction but a nuanced form of control and collaboration with Arvo.
-Through this exploration, we aim to demystify the enigma of Urbit and bridge the cosmic gap between Earth’s established software paradigms and the Martian vision of computational purity and simplicity.
+This paper begins with a brief explanation of Urbit's architecture, focusing on Arvo (Urbit's kernel) and Vere (Urbit's Runtime). 
+Following this, is an explanation of three core parts of the Urbit OS: 
+\texttt{\%khan}, \texttt{conn.c} and \texttt{\%lick}. 
+These serve as distinct conduits between the terrestrial and the Martian, facilitating not just interaction but a nuanced form of command and control of the entire system.
+Through this exploration, we aim to demystify the architecture of Urbit and bridge the gap between Earth’s established software paradigms and the Martian vision of computation.
 
@@
 
-Arvo's unique architecture avoids what is referred to as "event spaghetti" by maintaining a clear causal chain for every computation.
-Each chain begins with a Unix I/O event and progresses through a series of steps until the computation's terminal cause.
-This deterministic nature allows for a high level of predictability and control, a stark contrast to the non-deterministic nature of most Earth-based operating systems.
+Arvo's unique architecture avoids what is referred to as ``event spaghetti'' by maintaining a clear causal chain for every computation. 
+Each chain begins with a Unix I/O event and progresses through a series of steps until the computation either completes successfully or fails.
+This deterministic nature allows for a high level of predictability and control, a stark contrast to most Earth-based operating systems.
 
-The kernel's design as a "purely functional operating system" – or more accurately, an "operating function" – underscores its uniqueness.
+The kernel's design as a ``purely functional operating system'' – or more accurately, an ``operating function''" – underscores its uniqueness.
 Arvo operates on the principle that the current state is a pure function of its event log, a record of every action ever performed.
@@
 
-Arvo handles nondeterminism in an innovative way.
-The system, in essence, behaves like a stateful packet transceiver – events are processed, but there's no guarantee of completion, mirroring the behavior of packet dropping in networking.
-This approach to handling nondeterminism is unique and aligns with Arvo's overall philosophy of simplicity and determinism.
-
-Arvo's determinism is a key feature, stacking it atop a frozen instruction set known as Nock.
-This concept, though new to operating systems, is not foreign to computing.
-Similar to how CPU instruction sets like x86-64 are frozen at the chip level, Arvo freezes the instruction set at a higher level, enabling deterministic computation.
-This high-level determinism contrasts sharply with the non-determinism typically found in Earth's operating systems.
-
-Handling nondeterminism in Arvo is akin to a heuristic decision-making process similar to dropping a packet in networking.
-This approach views Arvo as a stateful packet transceiver, where the completion of events is not guaranteed, a stark difference from traditional computing models.
-Additionally, because Arvo runs on a VM, it can obtain nondeterministic information, such as stack traces from infinite loops, through the interpreter beneath it, further enhancing its operational capabilities.
-
-The vision of Urbit, with Arvo at its core, is to transition from developer-hosted web services on multiple foreign servers to self-hosted applications on a personal server.
-Each architectural decision in Arvo aligns with this vision, emphasizing user ownership and management of data.
-This approach sets it apart from the conventional model of cloud-based services and centralized data management prevalent in Earth-based systems.
+Arvo handles non-determinism in an innovative way. 
+The system, in essence, behaves like a stateful packet transceiver -- events are processed, but there's no guarantee of the event successfully completing.
+This mirrors the behavior of packet dropping in networking.
+This approach to handling non-determinism is unique and aligns with Arvo's overall philosophy of simplicity and determinism.
+Additionally, because Arvo runs on a VM (Vere), it can obtain non-deterministic information, such as stack traces from infinite loops, through the interpreter beneath it injecting events.
 
@@
 
-The architecture of Vere is bifurcated into two distinct parts: the \texttt{king} and the \texttt{serf}.
+The architecture of Vere is split into two distinct parts: the \texttt{king} and the \texttt{serf}.
 The \texttt{king} component is responsible for handling vane I/O and managing the effects on Earth.
@@
 
-A core opcode in Urbit's assembly language (\texttt{nock}) is the \texttt{hint} opcode (opcode 11) this allows Urbit to pass hints to Vere which produces effects on Earth.
-This bridges the Earth/Mars gap and allows for vanes to interact outside of Mars.
+A core opcode in Urbit's assembly language (Nock) is the \texttt{hint} opcode (opcode 11). 
+The \texttt{hint} opcode passes some value to the interpreter. 
+The interpreter can then choose what higher order action to take based on this value.
+For example, the simplest use of the \texttt{hint} opcode is the ordinary ``debug printf''.
+A more complex use of the opcode allows Urbit to pass hints to Vere which produces effects on Earth (send a network packet, get a keyboard input, etc.); bridging the Earth/Mars gap and allowing for vanes to interact outside of Mars.
 
@@
 This mechanism demonstrates a profound integration between Vere and Arvo, where Vere possesses direct knowledge of Arvo's state. 
-Notably, Vere can 'scry' or query Arvo's state without altering it, retrieving information as needed.
+Notably, Vere can ``scry'' or query Arvo's state without altering it, retrieving information as needed. 
 This capability allows Vere to maintain the integrity of Arvo's deterministic model while enabling dynamic interaction with the outside world.
 
-Urbit's use of Unix as its BIOS is a pivotal aspect of its runtime operation.
+Urbit's use of Unix as its BIOS is an important aspect of its runtime operation. 
 By leveraging the robust, widely-used Unix system, Vere ensures that Urbit can run on a broad range of hardware platforms, effectively making Urbit's Martian technology accessible and operable in Earth's diverse computing environments.
@@
 
-Khan, as a crucial component of Urbit's architecture, functions as a high-level thread interface, pivotal for managing both internal and external communications within the Urbit ecosystem.
-It is designed to efficiently handle complex I/O operations, employing an IO monad coupled with an exception monad.
-This dual-mechanism approach allows Khan to adeptly manage complex IO tasks while effectively handling potential failures.
+\texttt{\%khan}, as a crucial component of Urbit's architecture, functions as a high-level thread interface, pivotal for managing both internal and external communications within the Urbit ecosystem.
+It is designed to efficiently handle complex I/O operations, employing an I/O monad coupled with an exception monad.
+This approach allows \texttt{\%khan} to adeptly manage complex I/O tasks while effectively handling potential failures.
 
-Khan allows threads to be triggered from outside of Urbit.
-A thread is a monadic function that takes arguments and produces a result.
-It may perform input and output while running, so it is not a pure function.
-A thread's strength is that it can easily perform complex IO operations.
-It uses what's often called the IO monad (plus the exception monad) to provide a natural framework for IO.
-A thread's weakness is that it's impermanent and may fail unexpectedly.
+At its core, \texttt{\%khan}'s design revolves around the use of `threads', which are monadic functions designed to handle arguments and produce results.
+These threads are optimized for I/O operations, leveraging the I/O monad for structured and efficient task handling. 
+The exception monad further ensures robust management of potential failures and unexpected events.
+The weakness of threads is that it's impermanent and may fail unexpectedly. 
+Since it is usually relying on some external 
 In most of its intermediate states, it expects only a small number of events (usually one), so if it receives anything it didn't expect, it fails.
@@
 
-Khan's role in Urbit can be likened to that of a control plane.
-Its primary function is to execute threads via a Unix Socket and relay the results back.
-This setup supports various thread-running modes, enhancing the flexibility of Urbit's computing environment and allowing for diverse interaction methods with both internal and external processes.
+\texttt{\%khan} was conceived as a way to control Urbit ships from the exterior using threads.
+The concept evolved a fair bit from proposal to implementation. 
+In practice, \texttt{\%khan} is essentially an interface wrapper for Spider-based threads, which produces a somewhat strange (but not unprecedented) situation in which a vane relies on a piece of userspace infrastructure to function correctly.
+\texttt{\%spider} is an agent for transient thread-level operations.
 
-At its core, Khan's design revolves around the use of 'threads', which are monadic functions designed to handle arguments and produce results.
-These threads are optimized for I/O operations, leveraging the IO monad for structured and efficient task handling.
-The exception monad further ensures robust management of potential failures and unexpected events.
-
-Khan's implementation ensures that Urbit serves as a personal server, replacing multiple developer-hosted web services on foreign servers with self-hosted applications on a single personal server.
-Its design underscores Urbit's commitment to user ownership and data management, setting it apart from the conventional model of cloud-based services and centralized data management prevalent in Earth-based systems.
+\texttt{\%khan} allows for pre-written threads that allow for easy hosting and maintenance to be bundled and distributed; helping hosting companies and self-hosted ships to easily run efficiently.
 
@@
 \texttt{conn.c}\footnote{\url{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}} is a driver in Vere. 
-It is a part of the "King" (a.k.a.
-"Urth") process.
+It is a part of the ``King'' process. 
 It exposes a \texttt{Unix domain socket}\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}} at \texttt{/path/to/pier/.urb/conn.sock} for sending/receiving data from external processes.
 
-The functionality of conn.c, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of urbit's control plane.
-It serves as a way for unix processes to receive insights about what is happening on urbit.
-It also allows for pass commands either \%khan and the injection of raw kernel moves into the system.
-The implication of this means that whoever is running your urbit runtime controls your urbit.
+The functionality of \texttt{conn.c}, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of Urbit's control plane.
+It serves as a way for Unix processes to receive insights about what is happening on Urbit. 
+It also allows for pass commands either \texttt{\%khan} and the injection of raw kernel moves (events) into the system.
 
-From a technical perspective, conn.c accepts newt-encoded \texttt{++jam}med nouns which take the form \texttt{[request-id command arguments]}. 
+From a technical perspective, \texttt{conn.c} accepts newt-encoded \texttt{++jam}ed nouns (defined below) which take the form \texttt{[request-id command arguments]}. 
 The newt-encoded format is:
@@
 
+Where 
 \begin{itemize}
-  \item \texttt{V} is the version
-  \item \texttt{B} is the jam size in bytes (little endian)
-  \item \texttt{J} is the jammed noun (little endian)
+  \item \texttt{V} is the version (currently 0)
+  \item \texttt{B} is the size of the \texttt{++jam}ed noun in bytes (little endian)
+  \item \texttt{J} is the \texttt{++jam}ed noun (little endian)
 \end{itemize}
 
-This structure allows for a variety of commands to be executed, including
+This structure allows for a variety of commands to be executed, including:
 
@@
   \item \texttt{\%ovum} - the injection of raw kernel moves 
-  \item \texttt{\%fyrd} - a direct shortcut to Khan commands
+  \item \texttt{\%fyrd} - a direct shortcut to \texttt{\%khan} commands
   \item \texttt{\%urth} - runtime subcommands like \texttt{\%pack} or \texttt{\%meld}
@@
 
-A valid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[request-id output]}, where:
+A valid \texttt{conn.c} command produces a newt-encoded \texttt{++jam}ed noun with type \texttt{[request-id output]}, where:
 \begin{itemize}
@@
 
-Although also dealing with interprocess communication, \texttt{\%lick} \citep{UIP-101} was designed for a very different scenario than Khan:  to allow external processes, in particular hardware drivers, to intercommunicate with Urbit.
-(This breached the Earth/Mars divide.)  Thus \texttt{/sys/vane/lick} focuses on instrumenting a low-level noun interfaces over domain sockets.
+Although also dealing with interprocess communication, \texttt{\%lick} \citep{UIP-101} was designed for a very different use-case than \texttt{\%khan}: to allow external processes, in particular hardware drivers, to intercommunicate with Urbit.
+\texttt{\%lick} focuses on creating a generic noun interface over domain sockets.
 
@@
 Other vanes and applications ask \texttt{\%lick} to open an IPC port, notify it when something is connected or disconnected, and transfer data between itself and the Unix application.
-Lick works by opening a Unix socket for a particular process, which allows serialized IPC communications.
+\texttt{\%lick} works by opening a Unix socket for a particular process, which allows serialized IPC communications.
 These involve a jammed noun so the receiving process needs to know how to communicate in nouns.
-The IPC ports Lick creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of the \texttt{SOCK\_STREAM} type.
-The connexions are made via filepaths in \texttt{.urb/dev} of the pier.
+The IPC ports texttt{\%lick} creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of the \texttt{SOCK\_STREAM} type.
+The connections are made via file path in \texttt{.urb/dev} of the pier.
 
-The process on the host OS must therefore strip the first 5 bytes, \texttt{++cue}\footnote{\url{https://docs.urbit.org/language/hoon/reference/stdlib/2p#cue}} the jamfile, check the mark and (most likely) convert the noun into a native data structure.
+The process on the host OS must therefore strip the first 5 bytes, \texttt{++cue}\footnote{\url{https://docs.urbit.org/language/hoon/reference/stdlib/2p#cue}} the \texttt{++jam}ed noun, check the mark and (most likely) convert the noun into a native data structure.
 
@@
 
-The vane definition of \texttt{/sys/vane/lick} is even simpler than \texttt{/sys/vane/khan}:  it has no \texttt{++abet} core and primarily communicates to the \texttt{unix-duct} in its state.
+The vane definition of \texttt{\%lick} is even simpler than \texttt{\%khan}:  it has no \texttt{++abet} core and primarily communicates to the \texttt{unix-duct} in its state.
 The \texttt{+\$owner} is a \texttt{+\$duct} to handle the return \texttt{\%soak}.
@@
 Arvo will send three types of \texttt{\%soak}s to an open \texttt{\%lick} port. 
-These \texttt{\%soak}s are \texttt{\%connect} when the first connection is established on the Earth-side IPC port, \texttt{\%disconnect} when the last connection is broken on the Earth-side IPC port, and \texttt{\%error} when an error occurs.
+These \texttt{\%soak}s are \texttt{\%connect} when the first connection is established on the Earth-side IPC port,
+\texttt{\%disconnect} when the last connection is broken on the Earth-side IPC port,
+and \texttt{\%error} when an error occurs. 
 \texttt{\%lick} also will send a \texttt{\%disconnect} \texttt{\%soak} to every agent when Vere is started.
 
-Gall needs to wrap \texttt{\%soak} and \texttt{\%spit} to route properly.
-See e.g.
-\texttt{\%++ap-generic-take}.
-This lets multiple agents share sockets with the same name, and each agent can have its own folder.
+Gall needs to wrap \texttt{\%soak} and \texttt{\%spit} to route properly (see e.g. \texttt{\%++ap-generic-take}). 
+This lets multiple agents share sockets with the same name, while placing an agents sockets in a unique file path.
 
@@
 
-As Urbit had to deal with the reality of running on Earth software, Developers had to come up with various ways of communicating between these systems.
-Originally the two vanes \texttt{\%eyre} and \texttt{\%iris} were used as an HTTP client and server, breaking the ``Earth calls Mars; Mars doesn't call Earth'' dynamic.
+This exploration into Urbit's IPC interfaces, specifically \texttt{\%khan} and \texttt{\%lick}, has illuminated the profound architectural innovations that distinguish Urbit from traditional computing paradigms.
+By delving into the inner workings of Arvo and Vere, as well as the mechanisms facilitating communication between these components, we've uncovered a realm where simplicity and determinism reign supreme.
+This study not only demystifies the complex underpinnings of Urbit but also highlights its potential to redefine our interaction with digital ecosystems.
 
-Recently, two other vanes, \texttt{\%khan} and \texttt{\%lick} have been developed that hold true to the original idea of communication between Earth and Mars.
+Urbit's design philosophy, rooted in the principles of a more principled and coherent digital universe, challenges us to reconsider the foundations upon which our current software ecosystems are built.
+The contrast between the evolved, patchwork nature of Earth's software and the Maxwellian purity of Urbit's architecture serves as a poignant reminder of the potential for innovation in software design.
+As we stand at the crossroads of Earth's complexity and Mars' simplicity, the lessons drawn from Urbit's IPC interfaces beckon us towards a future where software is not only a tool but a reflection of a more streamlined and purposeful digital existence.
+
+The journey through Urbit's landscape offers a vision of what digital computation could become: a domain where efficiency, clarity, and integrity are not aspirational but foundational. 
+As we continue to navigate the evolving landscape of software development, the insights garnered from this analysis of Urbit will undoubtedly inspire further exploration and innovation. 
+In embracing the principles elucidated by Urbit's design, we pave the way for a future where the digital and human experiences are more intimately aligned, fostering a computing environment that is both revolutionary and deeply resonant with the principles of simplicity and purpose that guide our exploration of the digital frontier.
+
+%As Urbit had to deal with the reality of running on Earth software, developers had to come up with various ways of communicating between these systems.
+%Originally the two vanes \texttt{\%eyre} and \texttt{\%iris} were used as an HTTP client and server, breaking the ``Earth calls Mars; Mars doesn't call Earth'' dynamic. 
+%Recently, two other vanes, \texttt{\%khan} and \texttt{\%lick} have been developed that hold true to the necessity idea of communication between Earth and Mars.
 These vanes force Earth to speak to Mars in a language it can understand, nouns.

### 2024-03-11 mss.tex
Post

@@
 \usepackage{ustj}
+
+\addbibresource{mss.bib}
 
@@
               %% We toggle between these:
-              Manuscript submitted for review.\\
-              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
+              % Manuscript submitted for review.\\
+              {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
               ~ \\
@@
 \fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
-\fancyhead[RO]{What Hast Earth to do With Mars}
+\fancyhead[RO]{Urbit's IPC Interfaces}
 \fancyfoot[LE,RO]{\thepage}
@@
 %%MANUSCRIPT
-\title{What Hast Earth to do With Mars: An explanation of Urbit's IPC interfaces}
+\title{What Agreement Hath Mars With Earth?  An Exploration of Urbit's IPC Interfaces}
 \author{~\authorpatp \\ \affiliation}
@@
 \begin{abstract}
-In the evolving landscape of digital computation, the interaction between conventional software paradigms and innovative systems like Urbit presents a unique challenge.
-This paper delves into the architectural intricacies and interfaces of Urbit, contrasting its pristine, Maxwellian principles against the complex, evolved nature of Earth's software ecosystems.
-Through a comprehensive examination of Urbit’s kernel (Arvo) and runtime (Vere), alongside its high-level and low-level IPC interfaces (\texttt{\%khan} and \texttt{\%lick}, respectively), we elucidate the mechanisms that facilitate communication and control between these disparate computing realms.
-Our analysis aims to demystify Urbit's architecture and underscore its potential to bridge the gap between the familiar digital environments of Earth and the untapped possibilities of Martian computing.
-
-%  This paper explores the innovative approaches undertaken by developers to enable effective communication between the Urbit system (Mars) and conventional software systems (Earth).
-Initially, Urbit's integration with external systems relied on the use of two specific vanes, \texttt{\%eyre} and \texttt{\%iris}, serving as HTTP client and server respectively.
-This setup, however, presented challenges to the ideal of autonomous operation envisaged for Urbit, traditionally characterized by the principle that ``Earth always calls Mar; Mars never calls Earth.''
-
-%  Addressing these challenges, the paper introduces the development of two new vanes, \texttt{\%khan} and \texttt{\%lick}, which embody the core principle of uncontaminated communication.
-These advancements allow for a refined interaction model where Earth-based systems communicate with Urbit in a Urbit's own language, nouns.
-
-%The implications of these developments are significant, offering a novel framework for the integration of distinct computational ecosystems.
-This paper discusses the technical and conceptual underpinnings of these vanes, their development process, and the broader implications for Urbit's interaction with conventional software systems.
-
+\noindent \sloppy
+The idealistic Maxwellian principles of Urbit are somewhat at odds with the evolved nature of legacy software ecosystems.
+In order to facilitate practical computing, Urbit presents several interfaces for communication with external (“Earth”) systems.
+Through a comprehensive examination of Urbit's Arvo kernel and Vere runtime, alongside its high-level and low-level {\sc ipc} interfaces (\texttt{\%khan} and \texttt{\%lick}, respectively), we elucidate the architectural mechanisms that facilitate communication and control between these disparate computing realms.
+This exposition aims to demystify Urbit's architecture and underscore its potential to bridge the gap between the familiar digital environments of Earth and the untapped possibilities of Martian computing.
 \end{abstract}
@@
 
-In the realm of software, two worlds exist in stark contrast: 
-Earth, the familiar domain of established software paradigms, bustling with complex and varied digital ecosystems, 
-and Mars, the enigmatic realm of Urbit, a system as pristine and archaic as the Martian landscape itself. 
-This article will explore how Earth software establish a communication with the Martian code.
+In Urbit's conception of software, two worlds exist in stark contrast:  Earth, the familiar domain of established software paradigms, bustling with complex and varied digital ecosystems; 
+and Mars, the enigmatic realm of Urbit, a system as pristine and archaic as the Martian landscape itself.  This article will explore how Earth software establishes communication with Martian code.
 
-Unlike Earth, where software is a tapestry of evolution and patchwork, Mars presents a realm of purity. 
-It is a world where software exists as a Maxwellian construct, it is defined by a small set of axioms \citep{Nock4k}.
-Urbit code is untouched and unaltered by the non-Maxwellian intricacies that characterize Earth’s digital environments.
-Yet, in this isolation lies a profound awareness - an understanding by the Martians that there exist diverse forms of computing with which they must engage. 
-Mars, accommodates these external entities by allowing them to interact in the Martian language of nouns.
+\sloppy
+Unlike Earth, where code is a tapestry of evolution and patchwork, Mars aims to present a regime of functional-only idealism built on Nock, the ``Maxwell's laws of software''.
+Urbit envisions a world in which software exists as a Maxwellian construct defined by a small number of axioms \citep{Nock4K}.
+Urbit code is purely functional-as-in-language, and should be untouched and unaltered by the non-Maxwellian intricacies that characterize Earth’s digital environments.
+Underlying this sequestration lies a profound awareness by the ``Martians'' that there exist diverse forms of computing with which their system must nevertheless engage.
+Mars accommodates these external entities by allowing them to interact in the Martian \emph{lingua franca} of Nock nouns.
 
-In the original explanation of of Urbit, \citeauthor{Yarvin2010}, explicitly says 
-``The general structure of cross-planet computation is that Earth always calls Mars; Mars never calls Earth.
+\sloppy
+In the original public explanation of Urbit, \citeauthor{Yarvin2010} explicitly charges,
+\begin{quote}
+\sloppy
+The general structure of cross-planet computation is that Earth always calls Mars; Mars never calls Earth.
 The latter would be quite impossible, since Earth code is non-Maxwellian.
-There is only one way for a Maxwellian computer to run non-Maxwellian code: in a Maxwellian emulator.''
-This however has never been true for Urbit as it was implemented.
-Urbit has two vanes (best thought of as kernel modules) \texttt{\%eyre} and \texttt{\%iris} that are written as a specific interface to Earth (a web server and client, respectively).
-One can even make the argument that Urbit running on a virtual machine specifically breaks this convention.
+There is only one way for a Maxwellian computer to run non-Maxwellian code: in a Maxwellian emulator.
+\end{quote}
+\noindent
+However, this pristine discipline has never been strictly adhered to for any real Urbit implementation.
+Urbit has long provided two kernel-level vanes (which one may think of as microkernel modules) for interfacing with Earth's {\sc http} Web services, \texttt{\%eyre} (the server) and \texttt{\%iris} (the client).
+One can even make a reasonably strong argument that Urbit running as Nock on a virtual machine specifically breaks the convention to some extent.
 
-This paper begins with a brief explanation of Urbit's architecture, focusing on Arvo (Urbit's kernel) and Vere (Urbit's Runtime). 
-Following this, is an explanation of three core parts of the Urbit OS: 
-\texttt{\%khan}, \texttt{conn.c} and \texttt{\%lick}. 
+This article briefly explains Urbit's architecture, focusing on Arvo (Urbit's kernel) and Vere (Urbit's runtime VM).
+Following this, we explain three core parts of the Urbit OS interface:  \texttt{\%khan}, \texttt{conn.c} and \texttt{\%lick}.
 These serve as distinct conduits between the terrestrial and the Martian, facilitating not just interaction but a nuanced form of command and control of the entire system.
-Through this exploration, we aim to demystify the architecture of Urbit and bridge the gap between Earth’s established software paradigms and the Martian vision of computation.
+This exposition aims to demystify the architecture of Urbit and bridge the gap between Earth's established software paradigms and the Martian vision of computation.
 
@@
 \subsection{Arvo: The Heart of Martian Computing}
-Arvo, also known as Urbit OS, represents a paradigm shift in operating system design.
+
+The Arvo kernel (also referred to as ``Urbit OS'') represents a paradigm shift in operating system design.
 Unlike traditional operating systems that operate on preemptive multitasking and complex event networks, Arvo is a purely functional operating system, characterized by its deterministic nature and compact size.
-The entire Urbit stack is about 80,000 lines of code, with Arvo itself being only around 2,000 lines. 
+The entire Urbit stack is about 80,000 lines of code, with Arvo itself comprising around 2,000 lines.
 This small codebase is intentional, reflecting a philosophy that system administration complexity is directly proportional to code size.
 
-Arvo's unique architecture avoids what is referred to as ``event spaghetti'' by maintaining a clear causal chain for every computation. 
+Arvo's unique architecture avoids what is referred to as ``event spaghetti'' by maintaining a clear causal chain for every computation \citep{Whitepaper}.
 Each chain begins with a Unix I/O event and progresses through a series of steps until the computation either completes successfully or fails.
@@
 
-The kernel's design as a ``purely functional operating system'' – or more accurately, an ``operating function''" – underscores its uniqueness.
+The kernel's design as a ``purely functional operating system''—or, more accurately if idiosyncratically, an ``operating function''—underscores its uniqueness.
 Arvo operates on the principle that the current state is a pure function of its event log, a record of every action ever performed.
-This determinism is a significant deviation from the norm, where operating systems allow for programmatic alteration of global variables affecting other programs.
+This determinism is a significant deviation from the normal situation that obtains on Earth, wherein operating systems allow for arbitrary programmatic alteration of global variables affecting other programs.
 
-Arvo handles non-determinism in an innovative way. 
-The system, in essence, behaves like a stateful packet transceiver -- events are processed, but there's no guarantee of the event successfully completing.
-This mirrors the behavior of packet dropping in networking.
-This approach to handling non-determinism is unique and aligns with Arvo's overall philosophy of simplicity and determinism.
+Arvo innovatively handles non-determinism by refusing to log non-completed events.
+The system, in essence, behaves like a stateful packet transceiver—events are processed, but there is of course no guarantee of any particular event successfully completing.
+This pattern mirrors the behavior of packet dropping in networking.
+Arvo's approach to handling non-deter\-minism is unique and aligns with Urbit's overall philosophy of simplicity and determinism.
 Additionally, because Arvo runs on a VM (Vere), it can obtain non-deterministic information, such as stack traces from infinite loops, through the interpreter beneath it injecting events.
@@
 
-The architecture of Vere is split into two distinct parts: the \texttt{king} and the \texttt{serf}.
-The \texttt{king} component is responsible for handling vane I/O and managing the effects on Earth.
-In contrast, the \texttt{serf} operates as the process running the Arvo virtual machine (VM).
-This division of responsibilities allows for a clear delineation of tasks within the Vere runtime, enhancing its efficiency and robustness.
+Vere is architecturally split into two distinct parts:  the \texttt{king} (the Earth interface, more or less) and the \texttt{serf} (the Mars substrate for Nock and Arvo).
+The \texttt{king} component is responsible for handling vane I/O and managing the effects on and from Unix.
+The \texttt{serf} provisions the Arvo virtual machine (VM) and related services.
+This partition allows for a clear delineation of tasks within the Vere runtime, enhancing its efficiency and robustness.\footnote{The Ares project, in development, seeks to cleanly replace the \texttt{serf} process, a contingency enabled by this division of responsibility.}
 
-Communication between the \texttt{king} and \texttt{serf} is achieved through an Inter-Process Communication (IPC) port. 
-This IPC mechanism facilitates a continuous dialogue between the two components, ensuring that the \texttt{king} can effectively manage external interactions while the \texttt{serf} maintains the integrity of the Arvo VM.
+Communication between the \texttt{king} and \texttt{serf} is achieved via an inter-process communication ({\sc ipc}) port.
+This {\sc ipc} mechanism facilitates a continuous dialogue between the two components, ensuring that the \texttt{king} can effectively manage external interactions while the \texttt{serf} maintains the integrity of the Arvo VM.
 
-A core opcode in Urbit's assembly language (Nock) is the \texttt{hint} opcode (opcode 11). 
-The \texttt{hint} opcode passes some value to the interpreter. 
-The interpreter can then choose what higher order action to take based on this value.
-For example, the simplest use of the \texttt{hint} opcode is the ordinary ``debug printf''.
-A more complex use of the opcode allows Urbit to pass hints to Vere which produces effects on Earth (send a network packet, get a keyboard input, etc.); bridging the Earth/Mars gap and allowing for vanes to interact outside of Mars.
+Formally, Nock (and thus Urbit) is functional-as-in-language and free of side effects.
+In practice, the \texttt{hint} opcode 11 in Nock is used to pass some noun to the interpreter.
+The interpreter can then choose a higher order action to take based on this value, whether I/O, memory management, or other side effects.
+This allows for the \texttt{king} to pass hints to the \texttt{serf} which produces effects on Earth (send a network packet, get a keyboard input, etc.).
+This bridging of the Earth/Mars gap allows for vanes to interact outside of Mars.
 
-Key to Vere's functionality is how it processes events. 
+Event processing is key to Vere's functionality.
 Events from the Unix environment are passed to Vere, which then translates and injects them into Arvo.
@@
 
-Urbit's use of Unix as its BIOS is an important aspect of its runtime operation. 
+Urbit's runtime operation is effected by using Unix as its {\sc bios}.
 By leveraging the robust, widely-used Unix system, Vere ensures that Urbit can run on a broad range of hardware platforms, effectively making Urbit's Martian technology accessible and operable in Earth's diverse computing environments.
-This strategic use of Unix not only provides the necessary I/O and optimizations for Arvo but also ensures that Urbit's advanced and unique software architecture can function in tandem with the existing, established hardware and software ecosystems of Earth.
+This strategic use of Unix not only provides the necessary I/O and optimizations for Arvo but also ensures that Urbit's advanced and unique software architecture can function in tandem with the existing, established hardware and software eco\-systems of Earth.
 
@@
 This bidirectional communication channel ensures that Arvo can respond to external stimuli while remaining true to its functional and deterministic nature.
+Such events, from Arvo's perspective, are simply incidental injections to the event loop.
 
@@
 
-\section{\%khan - The High-Level Thread Interface of Urbit}
+\section{\texttt{\%khan}:  The High-Level Thread Interface}
 
-\texttt{\%khan}, as a crucial component of Urbit's architecture, functions as a high-level thread interface, pivotal for managing both internal and external communications within the Urbit ecosystem.
-It is designed to efficiently handle complex I/O operations, employing an I/O monad coupled with an exception monad.
+The \texttt{\%khan} vane functions as a high-level thread interface for managing both internal and external communications within the Urbit ecosystem.
+\texttt{\%khan} is designed to efficiently handle complex I/O operations, employing an I/O monad coupled with an exception monad.
 This approach allows \texttt{\%khan} to adeptly manage complex I/O tasks while effectively handling potential failures.
 
-At its core, \texttt{\%khan}'s design revolves around the use of `threads', which are monadic functions designed to handle arguments and produce results.
-These threads are optimized for I/O operations, leveraging the I/O monad for structured and efficient task handling. 
+\texttt{\%khan}'s core design revolves around the use of `threads', which are monadic functions designed to handle arguments and produce results.
+Threads are optimized for I/O operations, leveraging the I/O monad for structured and efficient task handling.
 The exception monad further ensures robust management of potential failures and unexpected events.
-The weakness of threads is that it's impermanent and may fail unexpectedly. 
-Since it is usually relying on some external 
-In most of its intermediate states, it expects only a small number of events (usually one), so if it receives anything it didn't expect, it fails.
-When code is upgraded, it's impossible to upgrade a running thread, so it fails.
+An Arvo thread is impermanent and may fail unexpectedly; it usually relies on some process external to Arvo with unknown existence or reliability.
+In most of its intermediate states, it expects only a small number of events (usually one), so if the thread receives anything it does not expect, it fails.
+Running threads are also not upgradable and fail across code upgrades.
+Threads are fragile but well-suited for their use case in managing such tenuous I/O operations.
 
-\texttt{\%khan} was conceived as a way to control Urbit ships from the exterior using threads.
-The concept evolved a fair bit from proposal to implementation. 
-In practice, \texttt{\%khan} is essentially an interface wrapper for Spider-based threads, which produces a somewhat strange (but not unprecedented) situation in which a vane relies on a piece of userspace infrastructure to function correctly.
-\texttt{\%spider} is an agent for transient thread-level operations.
-
-\texttt{\%khan} allows for pre-written threads that allow for easy hosting and maintenance to be bundled and distributed; helping hosting companies and self-hosted ships to easily run efficiently.
+\sloppy
+The \texttt{\%khan} vane was conceived as a way to control Urbit ships from the exterior using threads.\footnote{\texttt{\%khan} was originally called the ``control plane'' after the network routing distinction between the control plane and the data plane.}  \texttt{\%khan}'s conception evolved a fair bit from proposal to implementation.
+In practice, \texttt{\%khan} is essentially an interface wrapper for \texttt{\%spider}-based threads,\footnote{\texttt{\%spider} is an agent for managing transient thread-level operations, i.e.
+in \texttt{\%gall}.} which produces the topsy-turvy situation in which a vane relies on a piece of userspace infrastructure to function correctly.
+\texttt{\%khan} allows for pre-written threads that allow for easy hosting and maintenance to be bundled and distributed, helping both hosting companies and self-hosters to manage Urbit instances efficiently.
 
@@
 
-\texttt{conn.c}\footnote{\url{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}} is a driver in Vere. 
-It is a part of the ``King'' process. 
-It exposes a \texttt{Unix domain socket}\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}} at \texttt{/path/to/pier/.urb/conn.sock} for sending/receiving data from external processes.
+The \texttt{conn.c} driver in Vere\footnote{\href{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}{\texttt{vere/io/conn.c}}} forms part of the \texttt{king} process.
+It exposes a \href{https://en.wikipedia.org/wiki/Unix_domain_socket}{Unix domain socket} at \texttt{/path/to/pier/.urb/conn.sock} for sending and receiving data from external processes.
 
-The functionality of \texttt{conn.c}, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of Urbit's control plane.
+\texttt{conn.c}'s functionality, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of Urbit's control plane.
 It serves as a way for Unix processes to receive insights about what is happening on Urbit.
-It also allows for pass commands either \texttt{\%khan} and the injection of raw kernel moves (events) into the system.
+\texttt{conn.c} also allows for \texttt{\%khan} and Vere generally to inject raw kernel moves (events, if successful) into the system.
 
-From a technical perspective, \texttt{conn.c} accepts newt-encoded \texttt{++jam}ed nouns (defined below) which take the form \texttt{[request-id command arguments]}. 
-The newt-encoded format is:
+From a technical perspective, \texttt{conn.c} accepts newt-encoded \lstinline[style=inlinecode]{++jam}med nouns (defined below) which take the form
+\begin{lstlisting}[style=listingblock]
+[request-id command arguments]
+\end{lstlisting}
+\noindent
+Newt encoding is a way of communicating nouns as single data objects in an unambiguous format.
 
-\begin{verbatim}
+\begin{lstlisting}[style=listingblock]
 V.BBBB.JJJJ.JJJJ...
-\end{verbatim}
-
-Where 
-\begin{itemize}
-  \item \texttt{V} is the version (currently 0)
-  \item \texttt{B} is the size of the \texttt{++jam}ed noun in bytes (little endian)
-  \item \texttt{J} is the \texttt{++jam}ed noun (little endian)
-\end{itemize}
-
+\end{lstlisting}
+\noindent
+where \texttt{V} is the version (currently 0); \texttt{B} is the size of the \lstinline[style=inlinecode]{++jam}med noun in bytes (little-endian unary); and \texttt{J} is the \lstinline[style=inlinecode]{++jam}med noun (little-endian).
 This structure allows for a variety of commands to be executed, including:
@@
 \begin{enumerate}
-  \item \texttt{\%ovum} - the injection of raw kernel moves 
-  \item \texttt{\%fyrd} - a direct shortcut to \texttt{\%khan} commands
-  \item \texttt{\%urth} - runtime subcommands like \texttt{\%pack} or \texttt{\%meld}
-  \item \texttt{\%peek} - namespace scry requests into Arvo
-  \item \texttt{\%peel} - emulated namespace scry requests into Vere
+  \item \texttt{\%ovum}, the injection of raw kernel moves;
+  \item \texttt{\%fyrd}, a direct shortcut to \texttt{\%khan} commands;
+  \item \texttt{\%urth}, runtime subcommands like \texttt{\%pack} or \texttt{\%meld};
+  \item \texttt{\%peek}, namespace scry requests into Arvo; and
+  \item \texttt{\%peel}, emulated namespace scry requests into Vere.
 \end{enumerate}
+\noindent
+A valid \texttt{conn.c} command produces a newt-encoded \lstinline[style=inlinecode]{++jam}med noun with type \lstinline[style=inlinecode]{[request-id output]}, where:
+\begin{itemize}
+  \item \texttt{request-id} matches the input \texttt{request-id}; and
+  \item \texttt{output} depends on the \texttt{command}.
+\end{itemize}
+\noindent
+An invalid \texttt{conn.c} command produces a newt-encoded \lstinline[style=inlinecode]{++jam}med noun with type \texttt{[0 \%bail error-code error-string]}.
 
-A valid \texttt{conn.c} command produces a newt-encoded \texttt{++jam}ed noun with type \texttt{[request-id output]}, where:
-\begin{itemize}
-  \item \texttt{request-id} matches the input \texttt{request-id}
-  \item \texttt{output} depends on the \texttt{command}
-\end{itemize}
+\section[{\texttt{\%lick}:  The Low-Level {\sc ipc} Interface}]{\texttt{\%lick}:  The Low-Level IPC Interface}
 
-An invalid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[0 \%bail error-code error-string]}.
+Although also dealing with interprocess communication, the \texttt{\%lick} vane \citep{UIP-0101} was designed for a very different use-case than \texttt{\%khan}:  to allow external processes, in particular hardware drivers, to intercommunicate with Urbit.
+\texttt{\%lick} focuses on creating a generic noun interface over Unix domain sockets.
 
-\section{\%lick - The Low Level IPC Interface}
+\texttt{\%lick} manages {\sc ipc} ports and the communication between Urbit applications and {\sc posix} applications via these ports.
+Other vanes and applications request \texttt{\%lick} to open an {\sc ipc} port, notify it when something is connected or disconnected, and transfer data between itself and the Unix application.
+\texttt{\%lick} works by opening a Unix socket for a particular process, which allows serialized {\sc ipc} communications.
+These involve a \lstinline[style=inlinecode]{++jam}med noun so the receiving process needs to know how to communicate in nouns.
+The {\sc ipc} ports texttt{\%lick} creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of type \texttt{SOCK\_STREAM}.
+Connections are made via files in a directory under the pier, \texttt{.urb/dev}.
 
-Although also dealing with interprocess communication, \texttt{\%lick} \citep{UIP-101} was designed for a very different use-case than \texttt{\%khan}: to allow external processes, in particular hardware drivers, to intercommunicate with Urbit.
-\texttt{\%lick} focuses on creating a generic noun interface over domain sockets.
+The process on the host OS must therefore strip the first 5 bytes, \lstinline[style=inlinecode]{++cue} (unjam) the \lstinline[style=inlinecode]{++jam}med noun, check the mark, and (most likely) convert the noun into a native data structure.
 
-\texttt{\%lick} manages IPC ports, and the communication between Urbit applications and POSIX applications via these ports.
-Other vanes and applications ask \texttt{\%lick} to open an IPC port, notify it when something is connected or disconnected, and transfer data between itself and the Unix application.
-\texttt{\%lick} works by opening a Unix socket for a particular process, which allows serialized IPC communications.
-These involve a jammed noun so the receiving process needs to know how to communicate in nouns.
-The IPC ports texttt{\%lick} creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of the \texttt{SOCK\_STREAM} type.
-The connections are made via file path in \texttt{.urb/dev} of the pier.
+To understand what \texttt{\%lick} is doing, we need to briefly examine Unix's {\sc ipc} model.
+{\sc ipc} (``interprocess communication'') describes any way that two processes in an operating system's shared context have to communicate with each other.
+\texttt{\%lick} focuses on Unix domain sockets, which are merely file-like communication endpoints.
 
-The process on the host OS must therefore strip the first 5 bytes, \texttt{++cue}\footnote{\url{https://docs.urbit.org/language/hoon/reference/stdlib/2p#cue}} the \texttt{++jam}ed noun, check the mark and (most likely) convert the noun into a native data structure.
+For instance, a valid use of \texttt{\%lick} would use cards to communicate (Listing~\ref{lst:lick-cards}).  Outbound moves are specified as \lstinline[style=inlinecode]{%spit}.
 
-To understand what \texttt{\%lick} is doing, we need to look at Unix's IPC model briefly.
-IPC ("interprocess communication") describes any way that two processes in an operating system's shared context have to communicate with each other. 
-\texttt{\%lick} focuses on Unix domain sockets\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}}, which are just communication endpoints\footnote{\url{https://man7.org/linux/man-pages/man7/unix.7.html}}.
+\texttt{\%lick}'s vane definition is even simpler than \texttt{\%khan}'s:  it has no \texttt{++abet} nested-core pattern and primarily communicates to a \texttt{unix-duct} provided in its state.
+The \lstinline[style=inlinecode]{+$owner} is a \lstinline[style=inlinecode]{+$duct} to handle the return \lstinline[style=inlinecode]{%soak}.
 
-For instance, a valid use of \texttt{\%lick} would use cards that look like this:
+Arvo will send three types of \lstinline[style=inlinecode]{%soak}s to an open \texttt{\%lick} port.  These \lstinline[style=inlinecode]{%soak}s may be one of:
 
-\begin{verbatim}
+\begin{enumerate}
+  \item  \lstinline[style=inlinecode]{%connect} when the first connection is established on the Earth-side {\sc ipc} port;
+  \item  \lstinline[style=inlinecode]{%disconnect} when the last connection is broken on the Earth-side {\sc ipc} port; or
+  \item  \lstinline[style=inlinecode]{%error} when an error occurs.
+\end{enumerate}
+\noindent
+\texttt{\%lick} also will send a \lstinline[style=inlinecode]{%disconnect} \lstinline[style=inlinecode]{%soak} to every agent when Vere is started.
+
+Gall needs to wrap \lstinline[style=inlinecode]{%soak} and \lstinline[style=inlinecode]{%spit} to route properly (see e.g.
+\lstinline[style=inlinecode]{++ap-generic-take}).
+This permits multiple agents to share sockets with the same name while still organizing an agent's sockets by a unique path.
+
+\begin{lstlisting}[
+  float,
+  style=listingcode,
+  caption={Usage of \texttt{\%lick} cards.},
+  label={lst:lick-cards}
+]
 ++  init  [[%pass / %arvo %l %spin /control]~ this]
@@
         :_  this
-          [%pass /spit %arvo %l %spit /control %init area.state]~
+        :~  :*  %pass  /spit  %arvo  %l
+                %spit  /control  %init  area.state
+        ==  ==
       ==
@@
    ==
-\end{verbatim}
-
-The vane definition of \texttt{\%lick} is even simpler than \texttt{\%khan}:  it has no \texttt{++abet} core and primarily communicates to the \texttt{unix-duct} in its state.
-The \texttt{+\$owner} is a \texttt{+\$duct} to handle the return \texttt{\%soak}.
-
-Arvo will send three types of \texttt{\%soak}s to an open \texttt{\%lick} port. 
-These \texttt{\%soak}s are \texttt{\%connect} when the first connection is established on the Earth-side IPC port,
-\texttt{\%disconnect} when the last connection is broken on the Earth-side IPC port,
-and \texttt{\%error} when an error occurs. 
-\texttt{\%lick} also will send a \texttt{\%disconnect} \texttt{\%soak} to every agent when Vere is started.
-
-Gall needs to wrap \texttt{\%soak} and \texttt{\%spit} to route properly (see e.g. \texttt{\%++ap-generic-take}). 
-This lets multiple agents share sockets with the same name, while placing an agents sockets in a unique file path.
+\end{lstlisting}
 
@@
 
-This exploration into Urbit's IPC interfaces, specifically \texttt{\%khan} and \texttt{\%lick}, has illuminated the profound architectural innovations that distinguish Urbit from traditional computing paradigms.
-By delving into the inner workings of Arvo and Vere, as well as the mechanisms facilitating communication between these components, we've uncovered a realm where simplicity and determinism reign supreme.
-This study not only demystifies the complex underpinnings of Urbit but also highlights its potential to redefine our interaction with digital ecosystems.
+This exploration into Urbit's {\sc ipc} interfaces, in particular \texttt{\%khan}, \texttt{conn.c}, and \texttt{\%lick}, has sought to illuminate the architectural innovations that distinguish Urbit from traditional computing paradigms.
+These mechanisms permit Vere to balance Arvo's demand for pristine determinism against the host Unix OS's many competing software interfaces and tools.
+While Urbit itself obviously relies deeply on these, following this pattern in other systems could lead to a redefinition of how digital ecosystems interact.
+\tombstone{}
 
-Urbit's design philosophy, rooted in the principles of a more principled and coherent digital universe, challenges us to reconsider the foundations upon which our current software ecosystems are built.
-The contrast between the evolved, patchwork nature of Earth's software and the Maxwellian purity of Urbit's architecture serves as a poignant reminder of the potential for innovation in software design.
-As we stand at the crossroads of Earth's complexity and Mars' simplicity, the lessons drawn from Urbit's IPC interfaces beckon us towards a future where software is not only a tool but a reflection of a more streamlined and purposeful digital existence.
-
-The journey through Urbit's landscape offers a vision of what digital computation could become: a domain where efficiency, clarity, and integrity are not aspirational but foundational. 
-As we continue to navigate the evolving landscape of software development, the insights garnered from this analysis of Urbit will undoubtedly inspire further exploration and innovation. 
-In embracing the principles elucidated by Urbit's design, we pave the way for a future where the digital and human experiences are more intimately aligned, fostering a computing environment that is both revolutionary and deeply resonant with the principles of simplicity and purpose that guide our exploration of the digital frontier.
-
-%As Urbit had to deal with the reality of running on Earth software, developers had to come up with various ways of communicating between these systems.
-%Originally the two vanes \texttt{\%eyre} and \texttt{\%iris} were used as an HTTP client and server, breaking the ``Earth calls Mars; Mars doesn't call Earth'' dynamic. 
-%Recently, two other vanes, \texttt{\%khan} and \texttt{\%lick} have been developed that hold true to the necessity idea of communication between Earth and Mars.
-These vanes force Earth to speak to Mars in a language it can understand, nouns.
-
+\selectlanguage{USenglish}
 \printbibliography

### 2024-03-11 mss.tex
Break title lines better.
Post PDF update

@@
 %%MANUSCRIPT
-\title{What Agreement Hath Mars With Earth?  An Exploration of Urbit's IPC Interfaces}
+\title{What\,Agreement\,Hath\,Mars\,With\,Earth? \\ An Exploration of Urbit's IPC Interfaces}
 \author{~\authorpatp \\ \affiliation}
@@
 
+\begin{lstlisting}[
+  float,
+  style=listingcode,
+  caption={Usage of \texttt{\%lick} cards.},
+  label={lst:lick-cards}
+]
+++  init  [[%pass / %arvo %l %spin /control]~ this]
+++  on-arvo
+  |=  [=wire =sign-arvo]
+  ?+    sign-arvo  (on-arvo:def wire sign-arvo)
+      [%lick %soak *]
+    ?+    mark.sign-arvo  [~ this]
+        %connect
+      :_  this
+      :~  :*  %pass  /spit  %arvo  %l
+              %spit  /control  %init  area.state
+      ==  ==
+  ==  ==
+++  send-state
+  |=  =state
+  ^-  card:agent:gall
+  :*  %pass  /spit  %arvo  %l 
+      %spit  /control  %state
+      [slick:state face.state food.state live.state]
+  ==
+\end{lstlisting}
+
 \texttt{\%lick}'s vane definition is even simpler than \texttt{\%khan}'s:  it has no \texttt{++abet} nested-core pattern and primarily communicates to a \texttt{unix-duct} provided in its state.
@@
 
-\begin{lstlisting}[
-  float,
-  style=listingcode,
-  caption={Usage of \texttt{\%lick} cards.},
-  label={lst:lick-cards}
-]
-++  init  [[%pass / %arvo %l %spin /control]~ this]
-::
-++  on-arvo
-  |=  [=wire =sign-arvo]
-  ?+  sign-arvo  (on-arvo:def wire sign-arvo)
-      [%lick %soak *]
-      ?+    mark.sign-arvo  [~ this]
-        ::
-          %connect
-        ~&  >  "connect"
-        :_  this
-        :~  :*  %pass  /spit  %arvo  %l
-                %spit  /control  %init  area.state
-        ==  ==
-      ==
-  ==
-::
-++  send-state
-  |=  =state
-  ^-  card:agent:gall
-  :*  %pass 
-      /spit 
-      %arvo %l 
-      %spit 
-      /control 
-      %state 
-      [slick:state face.state food.state live.state]
-   ==
-\end{lstlisting}
-
 \section{Conclusion}

### 2024-03-11 mss.tex
Reword abstract.

@@
 \fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
-\fancyhead[RO]{Urbit's IPC Interfaces}
+\fancyhead[RO]{Urbit's {\sc ipc} Interfaces}
 \fancyfoot[LE,RO]{\thepage}
@@
 \noindent \sloppy
-The idealistic Maxwellian principles of Urbit are somewhat at odds with the evolved nature of legacy software ecosystems.
+The idealistic Maxwellian principles of a system built on Nock are somewhat at odds with the evolved and tangled nature of legacy software ecosystems.
 In order to facilitate practical computing, Urbit presents several interfaces for communication with external (“Earth”) systems.

### 2024-04-10 mss.tex
Proof

@@
 \newcommand{\authorpatp}{\patp{mopfel-winrux}}
-\newcommand{\affiliation}{Native Planet}
+\newcommand{\affiliation}{NativePlanet}
 
@@
               % Manuscript submitted for review.\\
-              {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
+              {\it Urbit Systems Technical Journal} I:1 (2024):  83–92. \\
               ~ \\
@@
 \pagenumbering{arabic}
-\setcounter{page}{1}
+\setcounter{page}{83}
 
@@
 Arvo's unique architecture avoids what is referred to as ``event spaghetti'' by maintaining a clear causal chain for every computation \citep{Whitepaper}.
-Each chain begins with a Unix I/O event and progresses through a series of steps until the computation either completes successfully or fails.
-This deterministic nature allows for a high level of predictability and control, a stark contrast to most Earth-based operating systems.
+Each chain begins with a Unix I/O event and progresses through a series of steps until the computation completes successfully or fails.
+This deterministic nature allows for a high degree of predictability, a stark contrast to most Earth-based operating systems.
 
@@
 Communication between the \texttt{king} and \texttt{serf} is achieved via an inter-process communication ({\sc ipc}) port.
-This {\sc ipc} mechanism facilitates a continuous dialogue between the two components, ensuring that the \texttt{king} can effectively manage external interactions while the \texttt{serf} maintains the integrity of the Arvo VM.
+This {\sc ipc} mechanism facilitates a dialogue between the two components, ensuring that the \texttt{king} can effectively manage external interactions while the \texttt{serf} maintains the integrity of the Arvo VM.
 
@@
 \texttt{\%khan} is designed to efficiently handle complex I/O operations, employing an I/O monad coupled with an exception monad.
-This approach allows \texttt{\%khan} to adeptly manage complex I/O tasks while effectively handling potential failures.
+This approach allows \texttt{\%khan} to manage complex I/O tasks while robustly handling potential failures.
 
@@
 The \texttt{\%khan} vane was conceived as a way to control Urbit ships from the exterior using threads.\footnote{\texttt{\%khan} was originally called the ``control plane'' after the network routing distinction between the control plane and the data plane.}  \texttt{\%khan}'s conception evolved a fair bit from proposal to implementation.
-In practice, \texttt{\%khan} is essentially an interface wrapper for \texttt{\%spider}-based threads,\footnote{\texttt{\%spider} is an agent for managing transient thread-level operations, i.e.
-in \texttt{\%gall}.} which produces the topsy-turvy situation in which a vane relies on a piece of userspace infrastructure to function correctly.
+In practice, \texttt{\%khan} is essentially an interface wrapper for \texttt{\%spider}-based threads,\footnote{\texttt{\%spider} is a \texttt{\%gall} agent to manage transient thread-level operations.} which produces the topsy-turvy situation in which a vane relies on a piece of userspace infrastructure to function correctly.
 \texttt{\%khan} allows for pre-written threads that allow for easy hosting and maintenance to be bundled and distributed, helping both hosting companies and self-hosters to manage Urbit instances efficiently.
@@
 The \texttt{conn.c} driver in Vere\footnote{\href{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}{\texttt{vere/io/conn.c}}} forms part of the \texttt{king} process.
-It exposes a \href{https://en.wikipedia.org/wiki/Unix_domain_socket}{Unix domain socket} at \texttt{/path/to/pier/.urb/conn.sock} for sending and receiving data from external processes.
+It exposes a \href{https://en.wikipedia.org/wiki/Unix_domain_socket}{Unix domain socket} at \texttt{/path/to/pier} \texttt{/.urb/conn.sock} for sending and receiving data from external processes.
 

### 2024-04-25 mss.tex
Final cut.

@@
 Formally, Nock (and thus Urbit) is functional-as-in-language and free of side effects.
-In practice, the \texttt{hint} opcode 11 in Nock is used to pass some noun to the interpreter.
+In practice, the \lstinline[style=inlinecode]{%hint} opcode 11 in Nock is used to pass some noun to the interpreter.
 The interpreter can then choose a higher order action to take based on this value, whether I/O, memory management, or other side effects.
-This allows for the \texttt{king} to pass hints to the \texttt{serf} which produces effects on Earth (send a network packet, get a keyboard input, etc.).
+This allows for the \texttt{king} to pass hints to the \texttt{serf} which in turn produces effects on Earth (sends a network packet, retrieves keyboard input, etc.).
 This bridging of the Earth/Mars gap allows for vanes to interact outside of Mars.
@@
 
-\section[{\texttt{\%lick}:  The Low-Level {\sc ipc} Interface}]{\texttt{\%lick}:  The Low-Level IPC Interface}
+\section[{\texttt{\%lick}:  The Low-Level \textsc{\textbf{ipc}} Interface}]{\texttt{\%lick}:  The Low-Level IPC Interface}
 
@@
 
-Gall needs to wrap \lstinline[style=inlinecode]{%soak} and \lstinline[style=inlinecode]{%spit} to route properly (see e.g.
-\lstinline[style=inlinecode]{++ap-generic-take}).
+Gall needs to wrap \lstinline[style=inlinecode]{%soak} and \lstinline[style=inlinecode]{%spit} to route properly (see e.\,g.\ \lstinline[style=inlinecode]{++ap-generic-take}).
 This permits multiple agents to share sockets with the same name while still organizing an agent's sockets by a unique path.
@@
 
-This exploration into Urbit's {\sc ipc} interfaces, in particular \texttt{\%khan}, \texttt{conn.c}, and \texttt{\%lick}, has sought to illuminate the architectural innovations that distinguish Urbit from traditional computing paradigms.
+\sloppy
+This exposition of Urbit's {\sc ipc} interfaces, in particular \texttt{\%khan}, \texttt{conn.c}, and \texttt{\%lick}, has sought to cast light upon architectural innovations that distinguish Urbit from traditional computing paradigms.
 These mechanisms permit Vere to balance Arvo's demand for pristine determinism against the host Unix OS's many competing software interfaces and tools.
